Efficiently Detecting Intrusions\\in the Internet of Things

### 2014-10-03 paper.tex
some touch-ups

Supporting multiple intrusion detection (ID) algorithms imposes significant
overhead on Internet of Things (IoT) devices. Fusion of the underlying
algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce id-foo, a framework for the
development of efficient intrusion detection systems (IDS). It consists of a
domain specific language (DSL) that allows describing ID algorithms in a formal
and functional way. The id-foo code generator then fuses these algorithms on a
functional level and produces more optimally organised source code. A
side-by-side comparison shows that id-foo-based code significantly reduces
memory footprint, message passing overhead and execution time in comparison to
a straightforward and sequential implementation of the same algorithms.

@@

The promise of the Internet of Things (IoT) is exciting. Soon my car will
notify the central heating system at home that I will be late, due to
unforeseen heavy traffic. It now can turn up the heat a bit later, thus
lowering my energy bills and adding to a greener world. Wonderful, I'm in.

But what if the energy company is capable of doing the same? What if they
decide to turn up the heat while I'm gone, making my home nice and cosy but
also cranking up my bill and their profit. No thank you, let me out.

Along with all the wonderful possibilities, the IoT also holds an enormous
threat: by opening our homes and lives to all these interconnected devices, we
also open them to everyone who is technologically savvy enough to break these
new virtual locks, only now most of the time without a trace.

In classic (wired) networks, members of the network are protected by a firewall
that cleanses all traffic before it actually enters the local network.
Firewalls are primarily focusing on the actual network traffic, prohibiting
access to certain destinations and/or services. Attackers though often use
complicated strategies that abuse flaws in the actual services themselves. The
firewall can't see these, they pass the wall of fire unnoticed and deliver
their payload to the system under attack.

At that point it becomes important to know that something went wrong and if
possible how it went wrong. This information will at least enable us to learn
from the incident and prepare for future situations or correctly assess the
impact and mitigate it. This is where intrusion detection steps in. An
intrusion detection (ID) probe monitors all traffic that passes the firewall
and looks for known patterns. To do this, it typically gets a copied feed of
all traffic from the firewall. Often the intrusion detection system (IDS) also
has a feedback link to the firewall, enabling it to lock out the evildoer based
on its origin, but most of the time post-mortem.

In a wireless network however, it is not possible to have this single point in
the network that can oversee all traffic. Every node is uniquely addressable,
without the need to pass by any other node. So if the security perimeter
diminishes and every device stands on its own, every device will have to
implement its own lines of defence.

There is one upside to this situation though: with protection and ID present on
every device, the services running on those devices can take advantage of these
security layers. If the IDS detects a problem, the service can immediately
terminate the communication with the attacker, before it actually experiences
the consequences. Just like a firewall, the localised IDS actually can be a
layer between the hostile outside world and the software running on the device.

This integration will be a key point in development of security layers for IoT
devices. More than in a classic network IDS, the algorithms that try to
discover the attacks operate with a large degree of uncertainty. They also
often have to collaborate with other nearby devices in the network, to unmask
the culprit. This collaboration also increases the level of uncertainty,
because no other device can be fully trusted. In the end the IDS on an IoT
device will be able to provide information about the trust it puts in other
nodes in the network, but it will be up to some application-level policy to
decide what to do with this information.

@@

Current developments in this area focus on programming frameworks to structure
the implementation of different algorithms \cite{todo}, but don't offer a way
to optimise the usage of resources. They also don't support systematic reuse to
create different configurations for different nodes or evolution in time.
\cite{todo}.

@@

The IoT is a very heterogeneous environment, both with respect to hardware and
software. Being able to address this in a transparent and automated way is a
key success factor for any kind of development in this area. For a
non-functional and generic layer, such as ID, this aspect becomes even more
paramount.

A solution that supports the integration of multiple algorithms and targets IoT
devices needs to consider the impact on the resources of these devices. These
devices not only need to survive using a limited battery, the production cost
should also taken into consideration. Different algorithms should be combined
as to avoid cumulating their impact.

@@

The remainder of this paper proceeds as follows: Section \ref{background}
details some of the ideas behind FCF and identifies the different aspects of
the problem space. Section \ref{design} describes the design we applied to
construct the framework, its language and code generator. Section
\ref{evaluation} evaluates an implementation and application of the domain
specific language (DSL) and code generator, using a set of ID algorithms found
in literature. Section \ref{related} explores related work in the field of ID
DSLs and code generation for the IoT. Finally, section \ref{conclusion}
summarises out findings, draws conclusions and identifies topics for future
work.

>>>>

Supporting multiple intrusion detection (ID) algorithms imposes significant
overhead on Internet of Things (IoT) devices. Fusion of the underlying
algorithms can alleviate this, yet, proves to be time-consuming, repetitive and
error-prone. To address this problem we introduce id-foo, a framework for the
development of efficient intrusion detection systems (IDS). It consists of a
domain specific language (DSL) that allows formally describing ID algorithms on
a functional level. The id-foo code generator then fuses these algorithms and
produces more optimally organised source code. A side-by-side comparison shows
that generated code reduces memory footprint, message passing overhead and
execution time in comparison to a straightforward sequential implementation of
the same algorithms.

@@

The promise of the Internet of Things (IoT) is exciting: my car will notify the
central heating system at my smart home \cite{aldrich2003smart} that I will be
late. It now can turn up the heat later, thus lowering my energy bills and
adding to a greener world. Wonderful, I'm in.

But what if the energy company can do the same? What if they are able to turn
up the heat while I'm gone, making my home nice and cosy but also cranking up
my bill and their profit. No thank you, let me out.

Along with all the wonderful possibilities, the IoT clearly holds an enormous
threat: by opening our homes and lives to all these interconnected devices, we
also open them to everyone who is able to break these new, virtual locks, often
without leaving a trace.

In classic networks, members are protected by a firewall that inspects all
traffic before it's passed on into the network. Firewalls focus on individual
network packets, prohibiting access to certain destinations and/or services.
But attackers often use strategies that abuse flaws in one or more services
themselves. The firewall can't distinguish these from legitimate packets, they
pass unnoticed and deliver their payload.

Now it becomes important to know that something went wrong and, if possible,
how it went wrong. This information will at least enable us to learn from the
incident and prepare for future situations or correctly assess the impact and
mitigate it. This is where intrusion detection steps in. An intrusion detection
(ID) probe monitors all traffic that passes the firewall and looks for known
patterns. To do this, it typically gets a copied feed of all traffic from the
firewall. Often the intrusion detection system (IDS) also has a feedback link
to the firewall, enabling it to lock out the evildoer based on its origin, but
most of the time this can't stop the initial attack.

In a mobile ad hoc network (MANET) however, it is not possible to have this
single point in the network that can oversee all traffic
\cite{mishra2004intrusion}. Every node is individually approachable, without
the need to pass by any other node. So if the security perimeter diminishes and
every device stands on its own, every device will have to implement its own
lines of defence.

There is one upside to this situation though: with protection and ID present on
every device, the services running on those devices can take advantage of these
security layers. If the IDS detects a problem, the service can immediately
terminate the communication with the attacker, before it actually experiences
the consequences. Just like a firewall, the IDS can be a layer between the
hostile network and the software on the device.

This integration will be a key point in development of security layers for IoT
devices. More than in a classic network IDS, the algorithms that try to
discover the attacks operate with a large degree of uncertainty. They also
often have to collaborate with other nearby devices in the network, to unmask
the culprit \cite{marchang2008collaborative,krontiris2009cooperative}. This
collaboration increases the level of uncertainty, because no other device can
be fully trusted. In the end the IDS on an IoT device will be able to provide
information about the trust \cite{ganeriwal2008reputation} it puts in other
nodes in the network, but it will be up to some application-level policy to
decide what to do with this information.

@@

Current developments in this area focus on programming frameworks to structure
the implementation of different algorithms \cite{valero2012di} and offer the
required basic functional components to implement actual algorithms
\cite{krontiris2008lidea}, but don't offer a way to optimise the usage of
resources. They also don't support systematic reuse to create different
configurations for different nodes or evolution in time.

@@

The IoT is a very heterogeneous environment, both with respect to hardware and
software. Being able to address this in a transparent and automated way is a
key success factor for any kind of development in this area. For a
non-functional layer, such as ID, this aspect becomes even more paramount.

A solution that supports the integration of multiple algorithms and targets IoT
devices needs to consider the impact on the resources of these devices. These
devices not only need to survive using a limited battery, the production cost
should also taken into consideration. Different algorithms should be merged to
avoid accumulating their impact.

@@

The remainder of this paper proceeds as follows: Section \ref{background}
details some of the ideas behind FCF and identifies the different aspects of
the problem space. Section \ref{design} describes the design we applied to
construct the framework, its domain specific language (DSL) and code generator.
Section \ref{evaluation} evaluates an implementation and application of the DSL
and code generator, using a set of ID algorithms found in literature. Section
\ref{related} explores related work in the field of ID DSLs and code generation
for the IoT. Finally, section \ref{conclusion} summarises out findings, draws
conclusions and identifies topics for future work.
